---
title: "Regression Analysis: Theory and Applications"
subtitle: "Spring 2025 - Dr. Alexander Fisher"
format:
   html: 
    page-layout: full
    toc: false
---

# Schedule

```{r load-libraries, echo = F, message = F, eval = F}
library(tidyverse)
library(knitr)
```

```{r smart-date-setup, echo = F, message = F, eval = F}
# set these:
firstDay = as.Date("2025-08-25")
firstDayID = "Mon"
lastDay = as.Date("2025-12-05")
lectureDays = c("Tue", "Thu")
labDays = c("Wed")
holidays = as.Date(c("2025-09-01", "2025-10-13", "2025-10-14", "2025-11-26",
                     "2025-11-27", "2025-11-28"))

### render table
allDates = seq(from = firstDay, to = lastDay, by = 'day')
schedule = data.frame(allDates,
                        dayOfWeek = weekdays(allDates, abbreviate=TRUE)) %>%
  filter(dayOfWeek %in% c(lectureDays, labDays))
schedule = schedule %>%
  mutate(lab = ifelse(dayOfWeek %in% labDays, TRUE, FALSE)) %>%
  mutate(Topic = ifelse(allDates %in% holidays, "NO CLASS", "")) %>%
  mutate(Date = format(allDates, format = "%b %d")) %>%
  mutate(Date = paste(dayOfWeek, Date)) %>%
  mutate(Week = rep("", nrow(schedule))) %>%
  mutate(Prepare = rep("", nrow(schedule))) %>%
  mutate(Notes = rep("", nrow(schedule))) %>%
  mutate(Assignment = rep("", nrow(schedule))) %>%
  mutate(Exam = rep("", nrow(schedule))) %>%
  mutate(Project = rep("", nrow(schedule))) %>%
  select(Week, Date, Topic, Prepare, Notes, Assignment)
# to be pasted into console then pasted into markdown:
schedule %>%
  kable(format = "markdown")
```

```{=html}
<style>
    .heatMap {
        width: 100%;
        text-align: center;
    }
    .heatMap th {
        word-wrap: break-word;
        text-align: center;
    }
<!--     .heatMap tr:nth-of-type(2n+1) { -->
<!--   background: #FFF; -->
<!-- } -->
    .heatMap tr:nth-of-type(3n-1) { background: #D3D3D3; }
</style>
```

<!-- useful icons ðŸ’»ðŸ—’ðŸ“ -->

::: heatMap
| Week | Date       | Topic                  | Prepare | Notes | Assignment |
|:-----|:-----------|:-----------------------|:--------|:------|:-----------|
| 1    | Tue Aug 26 | welcome to STA 221!    |         |       |            |
|      | Wed Aug 27 | lab 0: getting started |         |       |            |
|      | Thu Aug 28 |                        |         |       |            |
| 2    | Tue Sep 02 |                        |         |       |            |
|      | Wed Sep 03 |                        |         |       |            |
|      | Thu Sep 04 |                        |         |       |            |
| 3    | Tue Sep 09 |                        |         |       |            |
|      | Wed Sep 10 |                        |         |       |            |
|      | Thu Sep 11 |                        |         |       |            |
| 4    | Tue Sep 16 |                        |         |       |            |
|      | Wed Sep 17 |                        |         |       |            |
|      | Thu Sep 18 |                        |         |       |            |
| 5    | Tue Sep 23 |                        |         |       |            |
|      | Wed Sep 24 |                        |         |       |            |
|      | Thu Sep 25 |                        |         |       |            |
| 6    | Tue Sep 30 |                        |         |       |            |
|      | Wed Oct 01 |                        |         |       |            |
|      | Thu Oct 02 |                        |         |       |            |
| 7    | Tue Oct 07 |                        |         |       |            |
|      | Wed Oct 08 |                        |         |       |            |
|      | Thu Oct 09 |                        |         |       |            |
| 8    | Tue Oct 14 | NO CLASS               |         |       |            |
|      | Wed Oct 15 |                        |         |       |            |
|      | Thu Oct 16 |                        |         |       |            |
| 9    | Tue Oct 21 |                        |         |       |            |
|      | Wed Oct 22 |                        |         |       |            |
|      | Thu Oct 23 |                        |         |       |            |
| 10   | Tue Oct 28 |                        |         |       |            |
|      | Wed Oct 29 |                        |         |       |            |
|      | Thu Oct 30 |                        |         |       |            |
| 11   | Tue Nov 04 |                        |         |       |            |
|      | Wed Nov 05 |                        |         |       |            |
|      | Thu Nov 06 |                        |         |       |            |
| 12   | Tue Nov 11 |                        |         |       |            |
|      | Wed Nov 12 |                        |         |       |            |
|      | Thu Nov 13 |                        |         |       |            |
| 13   | Tue Nov 18 |                        |         |       |            |
|      | Wed Nov 19 |                        |         |       |            |
|      | Thu Nov 20 |                        |         |       |            |
| 14   | Tue Nov 25 |                        |         |       |            |
|      | Wed Nov 26 | NO CLASS               |         |       |            |
|      | Thu Nov 27 | NO CLASS               |         |       |            |
| 15   | Tue Dec 02 |                        |         |       |            |
|      | Wed Dec 03 |                        |         |       |            |
|      | Thu Dec 04 |                        |         |       |            |
:::
