---
title: "Lab 07: Review"
execute:
  freeze: auto
  echo: true
  eval: false
  warning: false
  message: false
format: html
bibliography: references.bib
---

::: callout-important
# Due date

This lab is not due for grade.
:::

## Conceptual Practice

### Exercise 

Review [ridge regression notes](/notes/lec16.html), [GLS notes](/notes/GLSintro.pdf), and [logistic regression notes](/notes/lec18.html). 

Summarize the main ideas, including (1) when the idea is appropriate, (2) the mathematical formulation, (3) how to fit the model to data.

## Additional applied practice

You will use the following packages in today’s lab. Add other packages as needed.

```{r}
#| label: load-packages
#| warning: false
#| message: false

library(tidyverse)
library(tidymodels)
library(knitr)
library(pROC)
```

### Data

```{r}
rice <- readr::read_csv("https://sta221-fa25.github.io/labs/data/rice.csv")
```


The data in this lab contains measures describing the shape and structure of two rice varieties - *Cammeo* and *Osmancik*. To curate the data set, researchers used images from over 3,000 grains a rice in these two varieties. They then used automated methods to process the image data and extract 7 morphological features (features related to the structure) from each image. The data were originally presented and analyzed in @cinar2019classification and was obtained from the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/545/rice+cammeo+and+osmancik).

This analysis will focus on the following variables:

-   `Class`: Cammeo or Osmancik

-   `Area`: Size of the rice grain measured in pixels

-   `Eccentricity`: A measure of how round the ellipse is, i.e., how close the shape of the grain is to a circle.

[Click here](https://archive.ics.uci.edu/dataset/545/rice+cammeo+and+osmancik) for the full data dictionary.

```{r}
#| lable: read-data
rice <- read_csv("data/rice.csv")
```

## Exercises

**Goal**: The goal of the analysis is to use Area and Eccentricity to identify grains from the Cammeo variety versus those from the Osmancik variety.

### Exercise 1

We'll begin by exploring the data. Create a scatterplot of `Area` versus `Eccentricity` with the color and shape of the points by `Class`.

### Exercise 2

Based on the plot from the previous exercise, do you think the two types of rice can be distinguished based on `Area` and `Eccentricity`. Briefly explain.

### Exercise 3

With this type of classification problem, it is common to test the performance of a logistic regression model by splitting the data into a *training* and a *test* set. This means that we will choose a random subset of the data to estimate the regression coefficient (training), and we will then test the predictions on the remaining observations (test).

-   Randomly select 75% of the rows of the data to create the training set `rice_train` . Use `set.seed(221)`.

-   Put the remaining 25% of the observations in the testing set called `rice_test`.

### Exercise 4

In a logistic regression model, the *log-odds* of the response being "1" (or a "success") is given by $$\log\Big(\frac{\pi_i}{1-\pi_i}\Big) = \mathbf{x}_i^\mathsf{T} \boldsymbol{\beta}$$

In this analysis a "success" is the Class "Osmancik".

-   What does each $\pi_i$ represent in the context of this analysis?

-   How is the probability of the response variable being "1" calculated from the log-odds? Show the mathematical steps to go from the log-odds to the probability in your response.

### Exercise 5

Use `rice_train` to fit the logistic regression model for the response variable `Class` on the predictors `Area` and `Eccentricity` .

-   Neatly display the output using 3 digits.
-   Does the intercept have any reasonable interpretation? If so, interpret the intercept. Otherwise explain why not.

### Exercise 6

-   Interpret the coefficients on `Area` and `Eccentricity` in the context of the data in terms of the odds and comment on whether they are useful predictors of `Class` in the model.

### Exercise 7

-   How would you expect the log-odds of the rice grain being of the Osmancik variety to change if the measures eccentricity changes from 0.85 to 0.9?

-   How would you expect the odds of the rice grain being of the Osmancik variety to change if the measures eccentricity changes from 0.85 to 0.9?

### Exercise 8

Now let's test our model on the test set. Use the `predict()` function to obtain the predicted probabilities of the rice being Osmancik for the observations in `rice_test.`

### Exercise 9

With these estimated probabilities, we can now try to classify the rice in the test set. Choose a threshold for assigning a class to each observation based on the estimated probability. Briefly explain your reasoning for selecting the threshold, including any analysis used to make your decision.

## Exercise 10

Compare the estimated class assignments you constructed with the actual classes. Comment on the result and thus the model performance.







