---
title: "Introduction to inference"
author: "Dr. Alexander Fisher"
format: html
---

\newcommand{\by}{\boldsymbol{y}}
\newcommand{\hby}{\hat{\boldsymbol{y}}}
\newcommand{\bone}{\boldsymbol{1}}
\newcommand{\bx}{\boldsymbol{x}}
\newcommand{\bX}{\boldsymbol{X}}
\newcommand{\be}{\boldsymbol{\varepsilon}}
\newcommand{\hbe}{\hat{\be}}

```{r}
#| label: load-packages
#| code-fold: true
#| code-summary: "View libraries and data sets used in these notes"
#| warning: false
#| message: false
library(tidyverse)
library(tidymodels)
library(DT)

pokemon <- read_csv("~/Downloads/pokemon_data.csv") # complete, population data
```

## The data 

As of 2025, there are 1025 pokemon in existence.

In all statistical inference tasks, we only have a *sample* from the population. Let's consider a random sample of the pokemon data, given below:

```{r}
#| echo: false
#| eval: false
for(i in 1:1000) {
  set.seed(i)
  sample = pokemon %>% 
  slice_sample(n = 15) %>%
    filter(name == "Pikachu") %>%
    nrow()
  if(sample != 0) {
    print(i)
  }
}
```


```{r}
set.seed(48) 
pokemon_sample <- pokemon |>
  slice_sample(n = 15) |>
  select(dexnum, name, height_m, weight_kg) |>
  arrange(dexnum)
  

datatable(pokemon_sample, rownames = FALSE, options = list(pageLength = 5),
           caption = "sample of 15 pokemon")
```

Let's investigate the question: do taller pokemon weigh more? 

::: panel-tabset
## plot

```{r}
#| echo: false
#| warning: false
pokemon_sample |>
  ggplot(aes(x = height_m, y = weight_kg)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE, color = 'steelblue') +
  theme_bw() +
  labs(title = "Pokemon weight vs height", x = "height (m)", y = "weight (kg)")
```

## linear model

```{r}
lm(weight_kg ~ height_m, data = pokemon_sample) |>
  tidy()
```

## plot code 

```{r}
#| eval: false
pokemon_sample |>
  ggplot(aes(x = height_m, y = weight_kg)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE, color = 'steelblue') +
  theme_bw() +
  labs(title = "Pokemon weight vs height", x = "height (m)", y = "weight (kg)")
```

:::




## Question

How can we tell if our estimates $\hat{\beta}$ are any good?

