---
title: "Correlation and intro to simple linear regresion"
author: "Dr. Alexander Fisher"
format: html
---

```{r}
#| echo: false
#| eval: false
weather = 
  read_csv("../data/rdu-weather-history-complete.csv") %>%
    filter(date >= as.Date("2021-09-01")) %>%
    filter(date <= as.Date("2021-09-30")) %>%
  select(date, tmin, tmax)

plot(weather$tmin, weather$tmax)

write_csv(weather, "../data/rdu-weather-history.csv")
```

```{r}
#| label: load-packages
#| code-fold: true
#| code-summary: "View libraries and data sets used in these notes"
#| warning: false
#| messge: false
library(tidyverse)
library(DT)
 
weather = 
  read_csv("https://sta221-fa25.github.io/data/rdu-weather-history.csv") %>%
  arrange(date)
    
```

## Learning objectives

By the end of the day you should be able to explain the following concepts:

1.  covariance
2.  correlation
3.  location and scale invariance
4.  ordinary least squares

## Example

<!-- The `cars` data set (from base R). The variable `speed` is recorded in miles per hour (mph) while stopping distance `dist` is recorded in feet. -->

This data set contains Raleigh Durham International Airport weather data pulled from the NOAA web service between September 01 and September 30, 2021. The data were sourced from [https://catalog.data.gov/](https://catalog.data.gov/dataset/local-weather-archive/resource/70f03146-c910-4baf-a1e3-9cc92bc7d9bb) August 28, 2025.

```{r}
#| echo: false
datatable(weather, rownames = FALSE, options = list(pageLength = 5),
           caption = "The min. and max. daily temperature at RDU international airport in Farenheit")
```

<!-- ```{r} -->

<!-- #| echo: false -->

<!-- datatable(cars, rownames = FALSE, options = list(pageLength = 5), -->

<!--           caption = "The speed of cars and the distances taken to stop. Note that the data were recorded in the 1920s.") -->

<!-- ``` -->

We've recorded 30 observations of two measurements. We are interested in the association between these two measurements.

We'll call the minimum daily temperature measurement "x" and the maximum daily temperature "y".

## Vocabulary

| Variable | Explanation |
|------------------|------------------------------------------------------|
| $y$ | The outcome variable. Also called "response" or "dependent variable". In prediction tasks, this is the variable we are interested in predicting. |
| $x$ | The predictor. Also called "covariate", "feature", or "independent variable". |

How are $x$ and $y$ associated?

::: panel-tabset
## Scatter plot

```{r}
#| echo: false
#| warning: false

weather %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point() +
  labs(title = "Minimum and maxmimum temperature at RDU in September, 2021") +
  theme_bw()
```

## Code

```{r}
#| eval: false

weather %>%
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point() +
  labs(title = "Minimum and maxmimum temperature (F) at RDU in September, 2021") +
  theme_bw()
```
:::

Notice: visualized this way, each of the thirty data points, $(x_i, y_i)$, is an element of two-dimensional space.

-   How would you describe the association?

::: callout-note
## Definition: covariance

$cov(x, y) = \frac{1}{n-1}\sum_{i =1}^n (x_i - \bar{x}) (y_i - \bar{y})$ $^*$

$^*$ Note that we divide by $n-1$ when computing the *sample covariance*

-   $\bar{x} = \frac{1}{n} \sum x_i$ (mean of x)
-   $\bar{y} = \frac{1}{n} \sum y_i$ (mean of y)
:::

```{r}
x = weather %>%
  select(tmin) %>%
  pull()
y = weather %>%
  select(tmax) %>%
  pull()

cov(x, y)

sum((x - mean(x)) * (y - mean(y))) / (30 - 1)
```

-   Should the association be the same if the thermometer recording measurements was consistently off by 2 degrees Farenheit?

```{r}
# compute covariance of true temperature measurement
weather %>%
  mutate(temp_min_true = tmin + 2,
         temp_max_true = tmax + 2) %>%
  summarize(cov(temp_min_true, temp_max_true)) %>%
  pull()
```

Covariance is **location invariant**.

-   Will the association stay the same if we recorded temperature in celsius?

```{r}
# compute the covariance between temperature in celsius
weather %>%
  mutate(temp_min_c = (tmin - 32) * 5 / 9,
         temp_max_c = (tmax - 32) * 5 / 9) %>%
  summarize(cov(temp_min_c, temp_max_c)) %>%
  pull()
```

Covariance is **not scale invariant**. That is, covariance does depend on the scale of the measurements.

## Correlation

::: callout-note
## Definition: correlation

$cor(x, y) = \frac{\sum_{i=1}^n (x_i - \bar{x}) (y_i - \bar{y})}{\left(\sum (x_i - \bar{x})^2 \sum (y_i - \bar{y})^2\right)^{1/2}}$

More concisely, we may write

$cor(x, y) = \frac{S_{xy}}{S_{xx}^{1/2} S_{yy}^{1/2}}$
:::

```{r}
# correlation in Farenheit vs Celsi
# compute the covariance between temperature in celsius
weather %>%
  mutate(temp_min_c = (tmin - 32) * 5 / 9,
         temp_max_c = (tmax - 32) * 5 / 9) %>%
  summarize(cor_F = cor(tmin, tmax), 
            cor_C = cor(temp_min_c, temp_max_c))
```

::: panel-tabset
## Exercise

Show $cor(x, y) = cor(ax + b, cy + d)$.

## Solution

Let $$
\begin{aligned}
x^* &= ax_i + b,\\
y^* &= cx_i + d,
\end{aligned}
$$

then we want to prove that $cor(x, y) = cor(x^*, y^*)$.

$$
\begin{aligned}
cor(x^*, y^*) &= \frac{\sum(ax_i + b - a\bar{x} - b)(cy_i + d - c\bar{y} - d)}{\sqrt{
\sum (ax_i + b - a\bar{x} - b)^2 \sum (cy_i + d - c\bar{y} - d)^2
}
}\\
&= \frac{ac \sum(x_i -\bar{x})(y_i - \bar{y})}{
ac \sqrt{\sum (ax_i + b - a\bar{x} - b)^2 \sum (cy_i + d - c\bar{y} - d)^2
}
}\\
&= \frac{
S_{xy}}{\sqrt{S_{xx} S_{yy}}
}
\end{aligned}
$$
:::

Correlation is **location and scale invariant**!

Additional facts about correlation:

::: callout-tip
## Fact 1

-   Correlation is not invariant to **monotone** transformations of the data
:::

::: callout-note
## Definition: monotone

$g$ is a **monotonic function** iff $x \leq y$ implies $g(x) \leq g(y)$
:::

::: panel-tabset
## Exercise

Show by example that correlation is not invariant to monotonic transformations.

## Solution

```{r}
set.seed(221)
x = c(1:10) 
logx = log(x)
y = rnorm(10)
cor(x, y)
cor(logx, y)
```
:::

::: callout-tip
## Fact 2

-   Correlation is *not* robust to outliers
:::

::: panel-tabset
## Example 1

```{r}
set.seed(221)
x = c(1:5)
y= x + rnorm(5)
plot(x, y)
cor(x, y)
```

## Example 2

```{r}
x2 = c(x, 0)
y2 = c(y, 20)
plot(x2, y2)
cor(x2, y2)
```
:::

::: callout-tip
## Fact 3

Correlation is bounded between -1 and 1.
:::

:::: panel-tabset
## Exercise

Show that $|cor(x, y)| \leq 1$

## Solution

Cauchy-Schwarz inequality:

Let $u$ and $v$ be vectors of dimension $n$, then

$$
|u^T v|^2 \leq (u^Tu) (v^Tv).
$$ To prove that $\|cor(x,y)\| \leq 1$, let $u = x - \bar{x}$ and let $v = y - \bar{y}$.

::: callout-important
Notice that the dimension of a vector inner product, e.g. $u^Tv$ is $1 \times 1$, in other words, it is a "scalar", a number.
:::
::::

::: callout-tip
## Fact 4

-   correlation is a measure of *linear* association
:::

::: panel-tabset
## Example 1

```{r}
x = -10:10
y = x^2
plot(x, y)
cor(x, y) 
```

## Example 2

```{r}
x = -10:10
y = x
plot(x, y)
cor(x, y)
```
:::

Since correlation is related to a linear relationship between the data, strong correlation implies that

$$
y_i = \beta_0 + \beta_1 x_i + \epsilon_i
$$

::: panel-tabset
## Exercise

Check out the interactive regression web app here:

<https://seeing-theory.brown.edu/regression-analysis/index.html#section1>
:::

## Nomenclature of simple linear regression

| Symbol | Description |
|------------|------------------------------------------------------------|
| $y$ | The response variable. Also called the "outcome" or "dependent variable". |
| $x$ | A covariate. Also called the "predictor", "feature" or "independent variable". |
| $\beta_0, \beta_1$ | These are population parameters, i.e. *fixed* and *unknown* constants. |
| $\hat{\beta}_0, \hat{\beta}_1$ | Estimates of $\beta_0, \beta_1$ based on a sample. |
| $\hat{y}$ | The prediction outcome. $\hat{y} = \hat{\beta_0} + \hat{\beta_1} x$. May also be referred to as the "fitted regression model". |
| $\epsilon$ | The error. Defined by the regression equation: $\epsilon = y - \beta_0 - \beta_1 x$. |
| $\hat{\epsilon}$ or $e$ | The residual, i.e. the difference between the outcome and the fitted model. Defined as $y - \hat{y}$, or equivalently, $y - \hat{\beta_0} - \hat{\beta_1}x$ |
