---
title: "Simple linear regression"
author: "Dr. Alexander Fisher"
format: html
---

```{r}
#| label: load-packages
#| code-fold: true
#| code-summary: "View libraries and data sets used in these notes"
#| warning: false
#| messge: false
library(tidyverse)
library(DT)
 
weather = read_csv("~/Downloads/rdu-weather-history.csv") %>%
    filter(date >= as.Date("2021-08-01")) %>%
    filter(date <= as.Date("2021-08-31"))

weather %>%
    
```

## Learning objectives

By the end of the day you should be able to explain the following concepts:

1.  covariance
2.  correlation
3.  location and scale invariance
4.  ordinary least squares

## Example

The `cars` data set (from base R). The variable `speed` is recorded in miles per hour (mph) while stopping distance `dist` is recorded in feet.

```{r}
#| echo: false
datatable(cars, rownames = FALSE, options = list(pageLength = 5),
          caption = "The speed of cars and the distances taken to stop. Note that the data were recorded in the 1920s.")
```

We've recorded 50 observations of two measurements. We are interested in the association between these two measurements.

We'll call the speed measurement "x" and the stopping distance "y".

## Vocabulary

| Variable  | Explanation |
|----|--------------|
|    $y$       | The outcome variable. Also called "response" or "dependent variable". In prediction tasks, this is the variable we are interested in predicting.    |
|  $x$         |  The predictor. Also called "covariate", "feature", or "independent variable".    |



Weather data: 

https://catalog.data.gov/dataset/local-weather-archive/resource/70f03146-c910-4baf-a1e3-9cc92bc7d9bb

This dataset contains Raleigh Durham International Airport weather data pulled from the NOAA web service 




